### 2019-09-30 index.Rmd
New webpage
See index.Rmd

---
title: "Water Utility Rates in California"
#author: "Team BA"
output: 
  html_document: 
    self_contained: no
    keep_md: yes
    theme: cosmo
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    code_folding: hide
    highlight: "pygments"
---

Welcome to a clearing house of water data. This page explores water utility prices, consumption, sources, and the number of water quality violations each utility has experienced.

To the left is a table of contents for the following page, but most importantly, immediately below is a map of water prices throughout California for the utilities that have greater than 3,000 customers. And for the data intrepid, at the bottom is the unerlying data consisting of the annual water production and deliveries reports, and rates.

```{r global, include=FALSE}

# Global
viridis_color <- "C"

# File paths
file_parent <- ""
file_paths <- list(
  code = paste0(file_parent, "Code/"),
  data = paste0(file_parent, "Data/"),
  data_owrs = paste0(file_parent, "Data/OWRS/Open-Water-Rate-Specification-master/full_utility_rates/California/")
); rm(file_parent)

###############
# Open Packages
###############
# General packages needed
library(stringr)
library(lubridate)
library(viridis)
# US Census Data
library(tidycensus)
library(maps)
library(sf)
library(leaflet)
library(ggridges)
# Utilities
library(ggfortify)
library(haven)
# Aesthetics
library(hrbrthemes)
library(gcookbook)
library(DT)
# Scraping
library(rvest)
# Analysis
library(glmnet)
library(lfe)
# Laod last to avoid masking functionality
library(tidyverse)

###############
# Open Data
###############
owrs_summary_full <- read_csv(paste0(file_paths$data, "OWRS/summary_table_cleaned.csv")) %>% 
  mutate(tier_number = as.character(tier_number))
ear_deliv <- read_csv(paste0(file_paths$data, "EAR/EAR 2013-2016 DELIVERIES FINAL 06-22-2018.csv"))
ear_prod <- read_csv(paste0(file_paths$data, "EAR/EAR 2013-2016 PRODUCTION FINAL 06-22-2018_1.csv"), 
                     col_types = cols(.default = "c")) %>% 
  bind_rows(read_csv(paste0(file_paths$data, "EAR/EAR 2013-2016 PRODUCTION FINAL 06-22-2018_2.csv"),
                     col_types = cols(.default = "c"))) %>% 
  type_convert()

## Geospatial Data
affected <- st_read(paste0(file_paths$data, "affected"), "affected")
util_sf <- st_read(paste0(file_paths$data, "Water_Districts"), "Water_Districts")

###############
# Clean Data
###############
# Clean the OWRS data
owrs_summary_full %<>% 
  arrange(pwsid) %>% 
  filter(!is.na(Tier_price)) %>% 
  mutate(Tier_volume = if_else(bill_type == "Tiered" & is.na(Tier_volume), Inf, Tier_volume))

# Simplify the OWRS data for the shapefile
owrs_summary <- owrs_summary_full %>%
  group_by(pwsid) %>% 
  mutate(tier_count = n()) %>% 
  ungroup() %>% 
  distinct(utility_name, pwsid, effective_date, bill_frequency, bill_type, 
           service_charge, commodity_charge, bill, tier_count)

# Clean EAR data
## Deliveries - i.e. use cases
ear_deliv <- ear_deliv %>%
  select(PWSID, Water.System.Classification, Year, Month, Date,
         `WATER DELIVERIES TO Single.family.Residential`, `WATER DELIVERIES TO  Multi.family.Residential`,
         `WATER DELIVERIES Total.Delivered Residential IN REVISED UNITS (Total Does not include Landscape Irrigation, Agricultural or to other PWS)`,
         `WATER DELIVERIES TO  Commercial.Institutional`:`WATER DELIVERIES TO  Industrial`,
         `WATER DELIVERIES TO  Agricultural`, `WATER DELIVERIES TO  Landscape.Irrigation`,
         `CALCULATED GPCD (Total delivery to residential in gallons per capita day)`) %>% 
  rename(classification = Water.System.Classification,
         deliv_res = `WATER DELIVERIES TO Single.family.Residential`,
         deliv_res_multi = `WATER DELIVERIES TO  Multi.family.Residential`,
         deliv_com_inst = `WATER DELIVERIES TO  Commercial.Institutional`,
         deliv_ind = `WATER DELIVERIES TO  Industrial`,
         deliv_irig = `WATER DELIVERIES TO  Landscape.Irrigation`,
         deliv_ag = `WATER DELIVERIES TO  Agricultural`,
         deliv_irig = `WATER DELIVERIES TO  Landscape.Irrigation`, 
         deliv_res_no_irig = `WATER DELIVERIES Total.Delivered Residential IN REVISED UNITS (Total Does not include Landscape Irrigation, Agricultural or to other PWS)`,
         deliv_gpcd = `CALCULATED GPCD (Total delivery to residential in gallons per capita day)`) %>% 
  mutate(deliv_res = select(., deliv_res, deliv_res_multi) %>% apply(1, sum, na.rm = T),
         Date = mdy(Date)) %>% 
  select(-deliv_res_multi) %>% 
  replace_na(list(deliv_res = 0, deliv_res_no_irig = 0, deliv_com_inst = 0, deliv_ind = 0,
                  deliv_ag = 0, deliv_irig = 0, deliv_gpcd = 0)) %>% 
  type_convert() %>% 
  mutate(deliv_total = select(., deliv_res, deliv_com_inst, deliv_ag) %>% apply(1, sum, na.rm = T)) %>% 
  mutate_at(vars(deliv_res:deliv_irig), .funs = list(perc = ~ . / deliv_total))

# Production - i.e. sources
ear_prod <- ear_prod %>%
  select(PWSID, Year, Month, Date,
         `WATER PRODUCED FROM GROUNDWATER`:`WATER SOLD TO ANOTHER PUBLIC WATER SYSTEM`,
         `CALCULATED GPCD (Total Potable Produced in gallons per capita day)`) %>% 
  rename(prod_ground = `WATER PRODUCED FROM GROUNDWATER`,
         prod_surface  =`WATER PRODUCED FROM SURFACE WATER`,
         prod_purch = `FINSIHIED WATER PURCHASED OR RECEIVED FROM ANOTHER PUBLIC WATER SYSTEM`,
         prod_sold = `WATER SOLD TO ANOTHER PUBLIC WATER SYSTEM`,
         prod_gpcd = `CALCULATED GPCD (Total Potable Produced in gallons per capita day)`) %>% 
  mutate(Date = mdy(Date)) %>% 
  mutate_at(vars(prod_ground:prod_gpcd), str_replace, "-", as.character(NA)) %>% 
  replace_na(list(prod_ground = 0, prod_surface = 0, prod_purch = 0,
                  prod_sold = 0, prod_gpcd = 0)) %>%
  mutate_at(vars(prod_ground:prod_gpcd), str_remove_all, ",") %>% 
  mutate_at(vars(prod_ground:prod_gpcd), str_remove_all, "\\.00") %>% 
  mutate_at(vars(prod_ground:prod_gpcd), as.numeric) %>% 
  mutate(prod_sold = (-prod_sold),
         prod_total = select(., prod_ground, prod_surface, prod_purch, prod_sold) %>% apply(1, sum, na.rm = T)) %>% 
  mutate_at(vars(prod_ground:prod_sold), .funs = list(perc = ~ . / prod_total))

###############
# Match/Join Data
###############
## Join EAR
ear <- ear_deliv %>% 
  full_join(ear_prod)

# Match Data
wat_ut_xwalk <- util_sf %>%
  as_tibble() %>% 
  select(-geometry) %>% 
  # Clean names to join
  mutate(AGENCYNAME = str_remove_all(AGENCYNAME, "District|Water|Irrigation"),
         AGENCYNAME = str_trim(AGENCYNAME, "both")) %>% 
  # Join
  fuzzyjoin::stringdist_left_join(
    owrs_summary %>% 
      # Clean names to joins
      mutate(utility_name = str_remove_all(utility_name, "District|Water|Irrigation|City Of"),
             utility_name = str_trim(utility_name, "both")) %>% 
      rename(AGENCYNAME = utility_name)
  ) %>% 
  filter(!is.na(pwsid))

# Join the utility data for the visualization
util_sf <- util_sf %>% 
  mutate(AGENCYNAME = str_to_title(AGENCYNAME)) %>% 
  # Add water utility data
  left_join(wat_ut_xwalk) %>% 
  filter(!is.na(pwsid)) %>% 
  # mutate(OBJECTID = as_factor(OBJECTID)) %>% 
  left_join(
    # Add violation counts
    affected %>% 
      select(OBJECTID, Freq) %>% 
      as_tibble() %>% 
      select(-geometry) %>% 
      mutate(OBJECTID = as.numeric(OBJECTID))
  ) %>% 
  mutate_at(vars(commodity_charge, service_charge, bill), round, 2)

# Make a total summary 
ear_summary <- ear %>% 
  mutate(quarter = quarter(Date)) %>% 
  filter(quarter %in% 1:4) %>% 
  group_by(PWSID, quarter) %>% 
  summarise_at(vars(deliv_res:prod_sold_perc), mean, na.rm = T) %>% 
  ungroup() %>% 
  gather(variable, value, -PWSID, -quarter) %>% 
  unite(temp, variable, quarter, sep = "_Q") %>% 
  spread(temp, value) %>% 
  left_join(
    ear %>% 
      group_by(PWSID) %>% 
      summarise_at(vars(deliv_res:prod_sold_perc), mean, na.rm = T) %>% 
      rename_at(vars(deliv_res:prod_sold_perc), funs(paste0(., "_annual")))
  ) %>% 
  rename(pwsid = PWSID) %>% 
  inner_join(wat_ut_xwalk) %>% 
  left_join(
    # Add violation counts
    affected %>% 
      select(OBJECTID, Freq) %>% 
      as_tibble() %>% 
      select(-geometry) %>% 
      mutate(OBJECTID = as.numeric(OBJECTID))
  )

# Add source and sales data to master data
util_sf <- util_sf %>% 
  left_join(
    ear_summary %>% 
      select(pwsid, prod_ground_perc_annual, prod_surface_perc_annual, prod_purch_perc_annual,
             prod_gpcd_annual) %>% 
      mutate_at(vars(prod_ground_perc_annual, prod_surface_perc_annual, prod_purch_perc_annual), ~ (. * 100) %>% round(0)) %>% 
      mutate(prod_gpcd_annual = prod_gpcd_annual %>% round(0))
  )

```

## California's Municipal Water Mapped

*Note*: The map is sparse because it only covers communities with over 3,000 customers.

```{r shiny_app, out.width="100%", echo = FALSE, warning = FALSE, message = FALSE, width = 400, height = 800}

# Creating labels
wat_ut_labels <- sprintf("<strong>%s</strong>
                          <br/>Bill Type: %s
                          <br/>Bill Frequency: %s
                          <br/>Number of Tiers: %s
                          <br/>Service Charge: %s
                          <br/>Consumption Charge: %s
                          <br/>Average Bill: %s
                          <br/>Ave. Consumption (Gal./Day Percapita): %s
                          <br/>Water Quality Violations: %s
                          <br/><strong>Source (Percent): %s</strong>
                          <br/>Groundwater: %s
                          <br/>Surface Water: %s
                          <br/>Purchased: %s",
                  util_sf$AGENCYNAME, util_sf$bill_type, util_sf$bill_frequency,
                  util_sf$tier_count, util_sf$service_charge, util_sf$commodity_charge,
                  util_sf$bill, util_sf$prod_gpcd_annual, util_sf$Freq, "", 
                  util_sf$prod_ground_perc_annual, util_sf$prod_surface_perc_annual, 
                  util_sf$prod_purch_perc_annual) %>%
  lapply(htmltools::HTML)

# Put the service charge in bins
## Service Charges
bins_serv <- seq(min(util_sf$service_charge), max(util_sf$service_charge), length.out = 10)
pal_serv <- colorBin("plasma", util_sf$service_charge, bins = bins_serv)
## Commondity
bins_com <- seq(min(util_sf$commodity_charge), max(util_sf$commodity_charge), length.out = 10)
pal_com <- colorBin("plasma", util_sf$commodity_charge, bins = bins_com)
## Total Bill
bins_bil <- seq(min(util_sf$bill), max(util_sf$bill), length.out = 10)
pal_bil <- colorBin("plasma", util_sf$bill, bins = bins_bil)

pal <- colorNumeric("plasma", domain = util_sf$bill)

# Plot
leaflet(util_sf, width = "100%") %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng = -122, lat = 37.9, zoom = 8) %>% 
  addPolygons(color = ~pal_bil(bill),
              fillOpacity = 0.5,
              weight = 1,
              label = wat_ut_labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"),
              group = "Average Bill") %>% 
  addPolygons(color = ~pal_serv(service_charge),
              fillOpacity = 0.5,
              weight = 1,
              label = wat_ut_labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"),
              group = "Service Charge") %>%
  addPolygons(color = ~pal_com(commodity_charge),
              fillOpacity = 0.5,
              weight = 1,
              label = wat_ut_labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"),
              group = "Water Price") %>%
  # addLegend(colors = ~pal_bil(bill),
  #           title = "Average Bill") %>%
  addLayersControl(
    overlayGroups = c("Average Bill", "Service Charge", "Water Price"),
    options = layersControlOptions(collapsed = FALSE),
    position = "bottomright"
  ) %>%
  hideGroup(c("Service Charge", "Water Price")) %>%
  leaflet.extras::addSearchOSM(options = list(position = "topright"))

```

## Water Prices

*Note*: Water price data only covers communities with over 3,000 customers. The average consumer is used to calculate the average service charge (fixed charge paid regardless of your consumption) and consumptive charge (price paid for consumption).

```{r sum_stats, echo = FALSE, warning = FALSE, message = FALSE, fig.height=4.5, fig.width=8, fig.align="center"}

# Distribution of Prices
owrs_summary %>% 
  select(bill, commodity_charge, service_charge) %>% 
  gather() %>% 
  mutate(key = case_when(
    key == "bill" ~ "Average Bill",
    key == "commodity_charge" ~ "Water Price",
    key == "service_charge" ~ "Service Charge",
    TRUE ~ key
  ),
  key = as_factor(key)) %>% 
  ggplot(aes(value, group = key, color= key)) + geom_line(stat = "density") + 
  scale_color_viridis_d(option = viridis_color, end = 0.8) +
  theme_ipsum() +
  labs(color = "Bill Type", x = "Price (USD)", y = "Density") +
  theme(legend.position = "bottom")

# Summary Statistics

owrs_summary_full %>% 
  filter(is.finite(Tier_volume)) %>% 
  # mutate(Tier_volume = if_else(is.infinite(Tier_volume), 100, Tier_volume)) %>% 
  ggplot(aes(x = Tier_volume, y = tier_number, fill = tier_number)) +
  geom_density_ridges(scale = 10, size = 0.25, rel_min_height = 0.03) +
  theme_ridges() +
  scale_x_continuous(limits=c(0, 150), expand = c(0.01, 0)) +
  #scale_y_discrete(expand = c(0.01, 0)) +
  scale_fill_brewer(palette = "YlOrRd")
  #scale_y_reverse(breaks=c(2000, 1980, 1960, 1940, 1920, 1900), expand = c(0.01, 0))

ggplot(owrs_summary_full, aes(x = Tier_price, y = tier_number, fill = tier_number)) +
  geom_density_ridges(scale = 10, size = 0.25, rel_min_height = 0.03) +
  theme_ridges() +
  scale_x_continuous(limits=c(0, 20), expand = c(0.01, 0)) +
  scale_fill_brewer(palette = "YlOrRd")
#scale_y_reverse(breaks=c(2000, 1980, 1960, 1940, 1920, 1900), expand = c(0.01, 0))

owrs_summary_full %>% 
  ggplot(aes(bill, approximate_median_income)) + 
  geom_point(size = 0.6) + 
  theme_ipsum() + 
  stat_smooth(method = "lm") +
  labs(x = "Average Bill (USD)", y = "Median Income (USD)") +
  xlim(c(0, 250))

```

## Extensions and Next Steps

* Defining affordability for water services:
  + A function of household income, number of people in HH, bill, regional average consumption ,and household consumption. Ultimately this is the percent of the income that a household spends on water conditional on regional consumption.
* Underlying details for each utility's produciton and deliveries data.
* Use the rate parser to build a measure of the average bill size.

### Downlaod Underlying Data

```{r data_download, echo=FALSE, warning = FALSE, message = FALSE}

# Names for the data set made public
util_sf_colnames <- c("Agency", "PWSID", "Date Updated", "Bill Frequency", "Bill Type",
                      "Average Fixed Charges", "Average Variable Change", "Average Total Bill",
                      "Number of Tiers at Agency", "Water Quality Violation Count",
                      "Source: Groundwater (%)", "Source: Surface Water (%)", "Source: Purchased (%)",
                      "Average Consumption Percapita per Day (Gallons)")

# Post the data
util_sf %>% 
  as_tibble() %>% 
  select(-geometry, -OBJECTID, -AGENCYUNIQ, -MODIFIEDBY, -SOURCECOMM, -Date_Data_, 
         -GlobalID, -AGENCYNAME.x, -AGENCYNAME.y, -SOURCE, -LASTMODIFI) %>% 
  datatable(., 
            extensions = 'Buttons', 
            colnames = util_sf_colnames,
            options = list(dom = 'Bfrtip', 
                           buttons = c('excel', "csv")))

```

### 2019-10-01 Code/affordability_metrics.R
move website to own directory for hosting via gcp

@@ -9,7 +9,7 @@ getwd()
 
 
 my_packages <- c("rgeos","raster","tidyverse","magrittr","haven","readxl","tools","usethis","RColorBrewer","ggrepel","sf","tictoc","lubridate","tidycensus")
-install.packages(my_packages, repos = "http://cran.rstudio.com")
+#install.packages(my_packages, repos = "http://cran.rstudio.com")
 lapply(my_packages, library, character.only = TRUE)
 
 
@@ -17,33 +17,91 @@ lapply(my_packages, library, character.only = TRUE)
 d <- read_csv("Data/OWRS/summary_table_cleaned_wide_formatv2.csv")
 View(d)
 
-# Calculate reasonable consumption: Let's say reasonable consumption is residential consumption in most efficient urban areas in CA 
-# e.g. San Francisco, which is 42 gpcd http://www2.pacinst.org/gpcd/map/#)
+# Calculate reasonable consumption: Let's say reasonable consumption is 50gpcd 
+# e.g. https://awwa.onlinelibrary.wiley.com/doi/full/10.5942/jawwa.2018.110.0002
 
 hh_sizes<-c(1:7) #household sizes in census go from 1 to 7
-gpcd<-42 # set gpcd for affordability threshold
+gpcd<-50 # set gpcd for affordability threshold
 
 d<-crossing(d,hh_sizes)
 d%<>%mutate(reasonable_consumption_kgal=(gpcd*hh_sizes*365/12)/1000)
-
-
-reasonable_consumption_kgal<-(gpcd*hh_sizes*365/12)/1000
-
+d$tier_0<-as.numeric(d$tier_0)
+d%<>%mutate(tier_0 = case_when(is.na(tier_0) ~ 0,
+                               TRUE ~ tier_0),
+            tier_1 = case_when(is.na(tier_1) ~ 10^5,
+                               TRUE ~ tier_1),
+            tier_2 = case_when(is.na(tier_2) ~ 10^6,
+                               TRUE ~ tier_2),                               
+            tier_3 = case_when(is.na(tier_3) ~ 10^7,
+                               TRUE ~ tier_3),
+            tier_4 = case_when(is.na(tier_4) ~ 10^8,
+                               TRUE ~ tier_4),                                           
+            tier_5 = case_when(is.na(tier_5) ~ 10^9,
+                               TRUE ~ tier_5),
[…]

### 2019-10-01 Code/affordability_metrics.R
some edits to affordability script comments

@@ -26,6 +26,8 @@ gpcd<-50 # set gpcd for affordability threshold
 d<-crossing(d,hh_sizes)
 d%<>%mutate(reasonable_consumption_kgal=(gpcd*hh_sizes*365/12)/1000)
 d$tier_0<-as.numeric(d$tier_0)
+
+#clean up water rates a bit, more, need to remove NAs from tier widths and tiers
 d%<>%mutate(tier_0 = case_when(is.na(tier_0) ~ 0,
                                TRUE ~ tier_0),
             tier_1 = case_when(is.na(tier_1) ~ 10^5,
@@ -88,10 +90,10 @@ return(total_bill)
   
 }
 
-d%<>%mutate(total_bill_by_hhsize=volRevCalc(d,d$reasonable_consumption_kgal))
-d%<>%mutate(mean_bill=volRevCalc(d,d$usage_ccf*0.748))
+d%<>%mutate(total_bill_by_hhsize=volRevCalc(d,d$reasonable_consumption_kgal),
+                                            mean_bill=volRevCalc(d,d$usage_ccf*0.748))
 
-
+d$approximate_median_income[which(d$median_income_category=="$150,000 to $199,999")]<-175000
 
 ##### AFFORDABILITY METRIC 1: % Median HH Income
 # The most commonly used affordability metric is
@@ -99,9 +101,11 @@ d%<>%mutate(mean_bill=volRevCalc(d,d$usage_ccf*0.748))
 d%<>%mutate(perc_MHI=100*12*mean_bill/approximate_median_income)
 
 
-##### AFFORDABILITY METRIC 2: Hours of Labor at Minimum Wage (MH)
+##### AFFORDABILITY METRIC 2: Hours of Labor at Minimum Wage (HM)
 # See https://awwa.onlinelibrary.wiley.com/doi/full/10.5942/jawwa.2018.110.0002
 
+mw = 12 #assume uniform minimum wage in CA at $12/hr. In future versions, would merge in relevant minimum wages
+d%<>%mutate(MH=total_bill_by_hhsize/mw)
 
-
-
+##### AFFORDABILITY METRIC 3: WARi 
+# See

### 2019-10-01 Code/cleaning_summary_table.R
updating service charge review
looked at the service charge bimonthly conversion again, updated based on review

@@ -26,8 +26,8 @@ unique(summary_table$bill_frequency)
 
 
 #la 09/30/2019 updated service charge based on comparison of OWRS raw file and summary_table.csv entry
-##for bi-monthly: Alameda County Water District 03/2018, Alhambra  City Of, Bakman Water Company, Garden Grove City, Westminister  City Of, 
-  #- but with Pomona  City of and Sweetwater Springs Water District - not sure what is going on there, 
+##for bi-monthly: Alameda County Water District 03/2018, Alhambra  City Of, Bakman Water Company, Garden Grove City, Westminister  City Of, Sweetwater Springs Water District 
+  #- but with Pomona  City of and Hayward City of - not sure what is going on there, 
 ##for quarterly: South Tahoe Public Utility District checks out fine
 ##for annual:Glenbrook Water Cooperative - not found in OWRS in github, not verified
 summary_table %<>%

### 2020-09-04 Untitled.Rmd
update work

---
title: "CA Affordabulity"
author: "Kyle Onda"
date: "9/3/2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidycensus)
library(sf)
library(beeswarm)
library(knitr)
library(stargazer)
```

## Profile of prevalence of private water utilities in CA (c. 2017))

There are ~ 2,800 Community Water Systems (defined as water systems serving at least 15 residential units on a year-round basis, generally applying to all municipal and institutional systems with residential components, but not commercial, school, and recreational facilities). Of these, 97.6% of the population is  served by the 653 systems that serve more than 3,300 people, while only 2.4% of the population is served by 2,173 small systems.

Overall, while 38% of community water systems are privately owned, they are disproportionately small systems, and so they only serve about 18% of the population. 94% of the population served by private utilities are served by utlities with more than 3,300 people. 
```{r boundaries, echo=FALSE,results='asis',message=FALSE, warning=FALSE}
b <- st_read("Data/boundaries.geojson") %>% filter(D_PWS_FED_=="C") 
b$size_cat= NA

b$size_cat[which(b$D_POPULATI<=500)] = "1. Very Small (<500)" 
b$size_cat[which(b$D_POPULATI>500)] = "2. Small (501-3300)" 
b$size_cat[which(b$D_POPULATI>3301)] = "3. Medium (3301-10,000)" 
b$size_cat[which(b$D_POPULATI>10001)] = "4. Large (10,001- 100,000)" 
b$size_cat[which(b$D_POPULATI>100000)] = "5. Very Large (>100,000)"

b$Ownership <-"Public"
b$Ownership[which(b$OWNER_TYPE=="P")]<-"Private"

size_summary <- b %>% group_by(size_cat)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry()

size_summary$perc_pop <- 100*size_summary$pop/sum(size_summary$pop)
size_summary <- arrange(size_summary,size_cat)
kable(size_summary, caption= "Size and population distribution of CA Water Utilities",digits=1)

private <- b%>%group_by(Ownership)%>%add_tally()%>%summarize(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() %>% ungroup()%>%mutate(perc_pop=100*pop/sum(pop))
kable(private, caption= "Population served by public and private utilities in CA",digits=1)

size_summary.p <- b %>% group_by(size_cat,Ownership)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry()

size_summary.p$perc_pop <- 100*size_summary.p$pop/sum(size_summary.p$pop)
size_summary.p <- arrange(size_summary.p,size_cat,Ownership)
kable(size_summary.p, caption= "Size and population distribution and ownership of CA Water Utilities",digits=1)

```

## Then parse the water rates

A common concern with private water utilities is that they would abuse their local monopolies and exploit market power to maximize profits and raise prices. It is difficult to ascertain the actual short and long-run costs of water utilities in the United States. It is also difficult to directly compare water prices across utilities, because there are complex rate structures such as increasing and decreasing block rates, seasonally varying rates, and water budget - based rates that vary with consumption levels. Moreover, in many places, the bulk of residential water use is spent on outdoor water uses such as lawn irrigation that are less salient to those concerned about affordability. Here, I use a survey of the rate structures of 387 (of which 79 are privately owned) of the largest CA utilities (comprising 30.5 million people, or 79% of the population served by water utilities in CA) in effect as of 2017 to calculate the average monthly water bill for households sized 1-7, assuming a reasonable water use for domestic purposes at 50 gallons per capita per day, in line with otehr scholars (M Teodoro 20xx, J Mumm 20XX, WHO 2011).

A  caveat here is that this subset of utilties for the most part represents phenomenon in larger California utilities. Cost structures and dynamics affecting water pricing and privatization in rural areas may differ substantially from those in urban areas. Below, we see that the vast majority of utilities with rates available for analysis have service populations greater than 1,000.

```{r rates pop,echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
rates <- read_csv("Data/affordability.csv")
priv <- select(b,PWSID,Ownership,D_POPULATI)
rates <- left_join(rates,b,by=c("pwsid"="PWSID")) 
rates <- rates[which(!is.na(rates$Ownership) & D_POPULATI>0),]
ggplot(rates, aes(x=log10(D_POPULATI), fill=Ownership)) +
  geom_density(alpha=0.4)
# Add mean lines

```

I then compare the monthly water bill (averaged to month for utilities with bimonthly, quarterly, and annual water bill regimes) for housheolds of size 1-7 between public and private utilities. In general, it does indeed seem to be the case that private utilties charge more on average for water than public utilities. Regression results show that on average, private utilties charge between \$3.50 and \$9.50  more per month depending on consumption levels than public utilies.

```{r rates, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
rates$Private <- 0
rates$Private[which(rates$Ownership=="Private")]<-1
ggplot(rates, aes(x=as.factor(hh_sizes), y=total_bill_by_hhsize, colour=Ownership)) +
  geom_boxplot()

m1<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==1),])
m2<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==2),])
m3<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==3),])
m4<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==4),])
m5<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==5),])
m6<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==6),])
m7<-lm(total_bill_by_hhsize~Private,data=rates[which(rates$hh_sizes==7),])
stargazer(m1,m2,m3,m4,m5,m6,m7,type='html')

```

However, might there be more complex patterns? For example, due to economies of scale, it's possible that larger utilities need to charge less for water. There does seem to be a negative correlation between utility size and water charges. Regressions indicate here that in general, across levels of consumption assuming households size 1-7, a doubling in service population is associated with a 5% decrease in monthly water bill (statistically significant <0.001).

```{r rates by size, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
ggplot(rates, aes(x=log(D_POPULATI+1), y=log(total_bill_by_hhsize), color=as.factor(hh_sizes), shape=as.factor(hh_sizes))) +
  geom_point() + 
  geom_smooth(method=lm)

```

Since, we saw above that private utilities tend to be smaller than public ones, we can attempt to control for this. Below, we see that controlling for population size, on average private utilities tend to have 11-16% higher water bills for a given household size at 50gpcd than public utilities. Note that this is an average effect, and that the distributions of water bills do overlap substantially between public and private utilties. In any case, this correlation holds when adding county fixed effects to account for regional clustering factors (such as many of the private utilities being clustered in the Los Angeles area)

```{r size private reg, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
m1<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==1),])
m2<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==2),])
m3<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==3),])
m4<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==4),])
m5<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==5),])
m6<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==6),])
m7<-lm(log(total_bill_by_hhsize)~log(D_POPULATI+1)+Private,data=rates[which(rates$hh_sizes==7),])
stargazer(m1,m2,m3,m4,m5,m6,m7,type='html')

```

## Census comparison

Look and see if there's any socioeconomic correlations

- Are richer places having higher prices?
- Are private utiltiies tending to be in richer or poorer places than public utilities?

If there are socioeconomic correlations, what are the implications for poorer people who are still in more expensive places? Need to review exisintg customer assistance programs/ water bill subsidy programs.

Having established that private utilities indeed tend to charge more for the same amount of water on average, we can investigate service levels. Is it plausible that the difference is entirely due to profit? Or are there differences in service levels or infrastructure investments being made?

## Infrastructure quality - leakage

One measure of performance quality might be Leakage. A water system with high levels of investment in asset management would be expected to have lower levels of leakage. CA in 2017 initiated mandatory water audits for most waterr utilities, that calculated, among many other indicators, 3 key leakage factors:

1. The Current Annual Real Losses (CARL) (an estimate of the amount of water physically leaking out of the system. That is, unlike NRW, this figure does not include unauthorized usage, unpaid water, and unmetered but consumed water). 
2. The Unavoidable Annual Real Losses (UARL) (an estimate of the lowest pracitcable level of leakage possible given the physical layout of the distribution system. All pipes and valves leak to some extent, and some systems naturally leak more than others due to their initial design and the nature of layout of the customer service connections.)

3. The Infrastructure Leakage Index (ILI), which is CARL/UARL.

```{r audi}
audit <- read_csv("Data/audit.csv")
```

## Water quality testing
```{r wq}
wq <- st_read("Data/ec_summary_aug2020/EC_Summary_Aug2020.shp")
wq2<- st_read("Data/hr2w_web_data_active-6-7-19.csv")
```

## And now census data

```{r census}

```

### 2020-09-15 Untitled.Rmd

@@
 date: "9/3/2020"
-output: html_document
+output:
+  html_document: default
+  word_document: default
 ---
@@
 ```
+
+## Data Sources
+
+### Census
+
+### OWRS Rate Information from CA Data Collaborative
+
+### Service Area Boundaries from CA DWR
+
+### Water Loss Audits from CA DWR
+
+### Electronic Annual Report from CA SWRCB
+
+#### Infrastructure repair costs
+
+#### Cut-offs
+
+#### Customer assistance programs
 
@@
 rates <- left_join(rates,b,by=c("pwsid"="PWSID")) 
-rates <- rates[which(!is.na(rates$Ownership) & D_POPULATI>0),]
+rates <- rates[which(!is.na(rates$Ownership)),]
 ggplot(rates, aes(x=log10(D_POPULATI), fill=Ownership)) +
@@
 
+## Shut-offs
+
+## Subsidy/ Cost assistance programs
+
 ## Census comparison
@@
 A water system with high levels of investment in asset management would be expected to have lower levels of leakage.
-CA in 2017 initiated mandatory water audits for most waterr utilities, that calculated, among many other indicators, 3 key leakage factors:
+The CA Water Resources Control Board in 2016 initiated mandatory water audits for most waterr utilities, that calculated, among many other indicators, 3 key leakage factors:
 
@@
 
-```{r audi}
+Below I compare the distribution of ILIs between public and private utilities where available for each year.
+It is clear that there is no significant difference in ILI between public and private utilities, so it is not clear if there are any performance gains to be had from privatization on this front.
+That said, the sample is limited to only 38 of the private utilities due to data reporting issues.
+
+```{r audi, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
 audit <- read_csv("Data/audit.csv")
+matches <- regmatches(audit$PWS_ID_OR_OTHER_ID, gregexpr("[[:digit:]]+", audit$PWS_ID_OR_OTHER_ID))
+audit$x<-matches
+audit$y<-audit$x
+audit$y[which(nchar(audit$x)==6)]<-paste0("0",audit$x[which(nchar(audit$x)==6)])
+audit$pwsid<-paste0("CA",audit$y)
+
+a<-select(audit,pwsid,PI_INFRASTRUCT_LEAKAGE_INDEX,REPORTING_YEAR) #%>% group_by(pwsid) %>%summarise(ili=mean(PI_INFRASTRUCT_LEAKAGE_INDEX))
+
+r <- left_join(rates,a,by=c("pwsid"))
+ggplot(r[which(r$hh_sizes==1),], aes(x=as.factor(REPORTING_YEAR), y=PI_INFRASTRUCT_LEAKAGE_INDEX, colour=Ownership)) +
+  geom_boxplot() 
 ```
 
-## Water quality testing
-```{r wq}
+## Safe Drinking Water Act compliance
+
+The Safe Drinking Water Act requires utilities to make a number of water quality tests and also file reports of the results of such tests to USEPA and/or state drinking water regulators.
+EPA determins on an annual basis whether a utility is a "serious violator", meaning they have some combination of repeated, unresolved major water quality tresting failures for regulated contaminants, or routinely fail to meet reporting requirements.
+I tabulate the number of years over the last 10 that a system has been in violation, tabulate by year.
+
+Among all the 2,826 systems for which compliance data is available 480 (17%) have been a "serious violator" for at least 1 year over the last decade (2011-2020).
+13.6% of public utilities and 25% of private utilities have been serious violators.
+However, when restricting to the 653 utilities with >3300 service populations, 53 (8%) have been serious violators.
+Of these larger utilities, 6 of 132 (4.5%) private utilities have been serious violatros while 47 of 468 (10%) of public utilities have been serious violators.
+
+We must be careful not to interpret this correlation causally. There may be a complex interplay between system size,
+
+```{r wq, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
 wq <- st_read("Data/ec_summary_aug2020/EC_Summary_Aug2020.shp")
-wq2<- st_read("Data/hr2w_web_data_active-6-7-19.csv")
+wq2<- read_csv("Data/hr2w_web_data_active-6-7-19.csv")
+v <- read_csv("Data/SDWA_SERIOUS_VIOLATORS.csv")
+v <- v%>% dplyr::select(PWSID,FISCAL_YEAR,SERIOUS_VIOLATOR,STATE) %>% filter(STATE=="CA")
+v$count=1
+v <- v%>%group_by(PWSID)%>%summarize(years_serious_violater=sum(count))
+
+bv<-left_join(b,v,by="PWSID") %>% st_drop_geometry()
+
+bv$years_serious_violater[which(is.na(bv$years_serious_violater))] <-0
+bv$serious_violator_last_decade=(bv$years_serious_violater>0)
+bv$count=1
+bv$Size_gr_3300=(bv$D_POPULATI>3300)
+bv$pop_affected=0
+bv$pop_affected[which(bv$serious_violator_last_decade==TRUE)]<-bv$D_POPULATI[which(bv$serious_violator_last_decade==TRUE)]
+
+k1<-bv%>%group_by(size_cat,Ownership)%>%summarise(utilities=sum(count),serious_violators=sum(serious_violator_last_decade),perc_utilitues=100*serious_violators/utilities,total_population=sum(D_POPULATI), population_affected=sum(pop_affected), perc_pop_affected=100*sum(pop_affected)/sum(D_POPULATI))
+
+k2<-bv%>%group_by(Size_gr_3300,Ownership)%>%summarise(utilities=sum(count),serious_violators=sum(serious_violator_last_decade),perc_utilitues=100*serious_violators/utilities,total_population=sum(D_POPULATI), population_affected=sum(pop_affected), perc_pop_affected=100*sum(pop_affected)/sum(D_POPULATI))
+
+kable(k1,digits=1)
+kable(k2,digits=1)
+
 ```
 
-## And now census data
-
-```{r census}
-
-```
+## Infrastructure repair and rehabilitation and replacement costs

### 2020-09-15 Untitled.Rmd
update ear data

@@
 # Add mean lines
+x <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip")
 

### 2020-09-15 Untitled.Rmd
Update Untitled.Rmd

--- 2020-09-15 Untitled.Rmd
@@
 # Add mean lines
+ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
+ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
+ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
+ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
+ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
+ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
 
+ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
+rm(ear13,ear14,ear15,ear16,ear17,ear18)
+save(ear,file="Data/EAR/ear.rds")
+
+hcf <- filter(ear,QuestionName=='WR 6 HCF DW Charges')
+x <- filter(ear, QuestionName=='')
 ```

### 2020-09-16 Untitled.Rmd
Update Untitled.Rmd

@@
 save(ear,file="Data/EAR/ear.rds")
+ear$QCat<-substr(ear$QuestionName,1,2)
 
-hcf <- filter(ear,QuestionName=='WR 6 HCF DW Charges')
-x <- filter(ear, QuestionName=='')
+wr <- filter(ear,QCat=="WR")
+
+hcf <- ear %>% 
+  filter(Year==2017 & (QuestionName=='WR 6 HCF DW Charges' |QuestionName=='WR 12 HCF DW Charges' )) %>% 
+  distinct(.keep_all=TRUE) %>%
+  mutate(ConsLevel=QuestionName,
+         MonthlyBill=as.numeric(QuestionResults)) %>%
+  select(PWSID,ConsLevel,MonthlyBill)
+
+#hcf12 <- filter(ear, QuestionName =="WR 12 HCF DW Charges")
+
 ```
@@
 ```{r rates, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
+r <- left_join(b,hcf,by="PWSID")
+wr_types <- wr %>% 
+  filter(QuestionName=="WR Type") %>% 
+  group_by(PWSID) %>% 
+  mutate(maxYear=max(Year)) %>% 
+  ungroup() %>% 
+  filter(Year==maxYear)
+
+no_rates <- wr_types %>% 
+  filter(QuestionResults=="-" | QuestionResults=="N/A") %>% 
+  mutate(rateType=QuestionResults) %>%
+  distinct(PWSID,.keep_all=TRUE)
+
+has_rates <- filter(wr, QuestionName=="WR Type") %>% distinct(PWSID)
+
+r <- anti_join(r,no_rates,by="PWSID")
+r <- semi_join(r,has_rates,by="PWSID")
+
+r$Private <- 0
+r$Private[which(r$Ownership=="Private")]<-1
+r$MonthlyBill[which(r$MonthlyBill>1000)]<-NA
+ggplot(filter(r, MonthlyBill>0), aes(x=as.factor(ConsLevel), y=MonthlyBill, colour=Ownership)) +
+  geom_boxplot()
+
+mr1<-lm(MonthlyBill~Private,data=r[which(r$ConsLevel=="WR 6 HCF DW Charges" & r$MonthlyBill>0),])
+mr2<-lm(MonthlyBill~Private,data=r[which(r$ConsLevel=="WR 12 HCF DW Charges"& r$MonthlyBill>0),])
+
+mr1.w<-lm(MonthlyBill~Private,data=r[which(r$ConsLevel=="WR 6 HCF DW Charges" & r$MonthlyBill>0),],weights=r$D_POPULATI[which(r$ConsLevel=="WR 6 HCF DW Charges" & r$MonthlyBill>0)])
+mr2.w<-lm(MonthlyBill~Private,data=r[which(r$ConsLevel=="WR 12 HCF DW Charges"& r$MonthlyBill>0),],weights=r$D_POPULATI[which(r$ConsLevel=="WR 12 HCF DW Charges"& r$MonthlyBill>0)])
+
 rates$Private <- 0
@@
 
+ggplot(filter(rates,hh_sizes==3), aes(x=log(D_POPULATI+1), y=log(total_bill_by_hhsize), color=as.factor(Private), shape=as.factor(Private))) +
+  geom_point()
+
+ggplot(filter(r,MonthlyBill>0 & ConsLevel=="WR 6 HCF DW Charges" & D_POPULATI>10000), aes(x=log(D_POPULATI), y=log(MonthlyBill), color=as.factor(Private))) +
+  geom_point() + 
+  geom_smooth(method=lm)
+
 ```
@@
 
+r1<-lm(log(MonthlyBill)~log(D_POPULATI+1)+Private,data=r[which(r$ConsLevel=="WR 6 HCF DW Charges" & r$MonthlyBill>0 & r$D_POPULATI>3300),])
+r2<-lm(log(MonthlyBill)~log(D_POPULATI+1)+Private,data=r[which(r$ConsLevel=="WR 12 HCF DW Charges" & r$MonthlyBill>0 & r$D_POPULATI>3300),])
+
+stargazer(r1,r2,type='html')
+
 ```
@@
 ## Shut-offs
+```{r shutoffs}
+conn <- ear %>% filter(SectionName=="Connections" & Year==2018)
+conn_sfr <- conn %>% filter(QuestionName=="SFR Potable Total" & Year==2018)
+conn_mfr <- conn %>% filter(QuestionName=="MFR Potable Total" & Year==2018)
+conn_total <- conn %>% filter(QuestionName=="T Potable Total" & Year==2018)
+
+shutoffs <- ear %>% filter(QuestionName=="")
+```
 

### 2020-09-16 paper.Rmd
update

--- 2020-09-16 Untitled.Rmd
@@
 conn_mfr <- conn %>% filter(QuestionName=="MFR Potable Total" & Year==2018)
-conn_total <- conn %>% filter(QuestionName=="T Potable Total" & Year==2018)
+conn_total <- conn %>% filter(QuestionName=="T Potable Total" & Year==2018) %>%rename(connections=QuestionResults) %>% mutate(connections=as.numeric(connections)) %>% select(PWSID,connections) %>% distinct(.keep_all=TRUE)
 
-shutoffs <- ear %>% filter(QuestionName=="")
+shutoffs <- ear %>% filter(substr(QuestionName,1,12)=="WR SHUT_OFFS" & Year==2018)
+
+pws_shutoff_collect <- shutoffs %>% filter(QuestionName=="WR SHUT_OFFS Once No Collect" & is.na(QuestionResults))
+
+shutoffs_total <- shutoffs %>%
+  filter(QuestionName=="WR SHUT_OFFS Once SF Total" |
+           QuestionName=="WR SHUT_OFFS More Than Once SF Total" |
+           QuestionName=="WR SHUT_OFFS More Than Once MF Total" |
+           QuestionName=="WR SHUT_OFFS Once SF Total") %>%
+  group_by(PWSID) %>%
+  summarise(shutoffs=sum(as.numeric(QuestionResults),na.rm=TRUE))%>%
+  select(PWSID,shutoffs) %>%
+  distinct(.keep_all=TRUE) %>% semi_join(pws_shutoff_collect,by="PWSID")
+
+shutoff_comments <- shutoffs %>% filter(QuestionName=="WR SHUT_OFFS Comments")
+  re
+shutoff_pa<- ear%>%filter(QuestionName=="WR SHUT_OFFS Payment Assistance Total" & Year==2018) %>%distinct(PWSID, QuestionResults) %>% rename(shutoff_pa=QuestionResults) %>% filter(!is.na(shutoff_pa)) %>% mutate(shutoff_pa=as.numeric(shutoff_pa))
+
+conn_shutoff <- left_join(conn_total,shutoffs_total,by="PWSID") %>% mutate(shutoff_rate=shutoffs/connections) %>% filter(connections>0) %>% filter(shutoff_rate<=1)
+
+r_shutoff <- left_join(r,conn_shutoff,by="PWSID") %>% distinct(PWSID,connections,.keep_all=TRUE) %>%
+ggplot(filter(r_shutoff), aes(x=size_cat, y=shutoff_rate, colour=Ownership)) +
+  geom_boxplot()
+
+lm()
+
 ```

### 2020-09-17 paper.Rmd
update

@@
 # Add mean lines
-ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
-ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
-ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
-ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
-ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
-ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
-
-ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
-rm(ear13,ear14,ear15,ear16,ear17,ear18)
-save(ear,file="Data/EAR/ear.rds")
+# ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
+# ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
+# ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
+# ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
+# ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
+# ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
+# 
+# ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
+# rm(ear13,ear14,ear15,ear16,ear17,ear18)
+# save(ear,file="Data/EAR/ear.rds")
+load("Data/EAR/ear.rds")
 ear$QCat<-substr(ear$QuestionName,1,2)
@@
 ## Shut-offs
-```{r shutoffs}
+```{r shutoffs, echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
 conn <- ear %>% filter(SectionName=="Connections" & Year==2018)
@@
 conn_mfr <- conn %>% filter(QuestionName=="MFR Potable Total" & Year==2018)
-conn_total <- conn %>% filter(QuestionName=="T Potable Total" & Year==2018) %>%rename(connections=QuestionResults) %>% mutate(connections=as.numeric(connections)) %>% select(PWSID,connections) %>% distinct(.keep_all=TRUE)
+conn_total <- conn %>% filter(QuestionName=="T Potable Total" & Year==2018) %>% rename(connections=QuestionResults) %>% mutate(connections=as.numeric(connections)) %>% select(PWSID,connections) %>% distinct(.keep_all=TRUE)
 
@@
 shutoff_comments <- shutoffs %>% filter(QuestionName=="WR SHUT_OFFS Comments")
-  re
+
 shutoff_pa<- ear%>%filter(QuestionName=="WR SHUT_OFFS Payment Assistance Total" & Year==2018) %>%distinct(PWSID, QuestionResults) %>% rename(shutoff_pa=QuestionResults) %>% filter(!is.na(shutoff_pa)) %>% mutate(shutoff_pa=as.numeric(shutoff_pa))
@@
 
-r_shutoff <- left_join(r,conn_shutoff,by="PWSID") %>% distinct(PWSID,connections,.keep_all=TRUE) %>%
-ggplot(filter(r_shutoff), aes(x=size_cat, y=shutoff_rate, colour=Ownership)) +
+r_shutoff <- left_join(r,conn_shutoff,by="PWSID") %>% distinct(PWSID,connections,.keep_all=TRUE) 
+ggplot(r_shutoff, aes(x=size_cat, y=shutoff_rate, colour=Ownership)) +
   geom_boxplot()
 
-lm()
+r_shutoff_model <- lm(
+  shutoff_rate ~ Private + log(D_POPULATI),
+  data = filter(r_shutoff, D_POPULATI>3300)
+#  family = quasibinomial
+)
 
+summary(r_shutoff_model)
 ```
@@
 ## Subsidy/ Cost assistance programs
+```{r subsidies}
+subs <- ear %>% filter(substr(QuestionName,1,5)=="WR LL" & Year==2018)
+subsidies <- ear %>% 
+  filter(QuestionName=="WR LL Subsidies" & Year==2018) %>% 
+  #filter(QuestionResults!="-") %>% 
+  select(PWSID,QuestionResults) %>% 
+  rename(subsidy=QuestionResults) %>% distinct(.keep_all=TRUE)
+
+subsidies_LI <- subs %>% 
+  filter(QuestionName=="WR LL Eligibility-Low Income" & Year==2018) %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(LowIncome=QuestionResults) %>% distinct(.keep_all=TRUE) %>% mutate(LowIncome="Yes")
+
+subsidies_D <- subs %>% 
+  filter(QuestionName=="WR LL Eligibility-Disabled" & Year==2018) %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(Disabled=QuestionResults) %>% distinct(.keep_all=TRUE)  %>% mutate(Disabled="Yes")
+
+subsidies_S <- subs %>% 
+  filter(QuestionName=="WR LL Eligibility-Seniors" & Year==2018) %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(Seniors=QuestionResults) %>% distinct(.keep_all=TRUE) %>% mutate(Seniors="Yes")
+
+subsidies_M <- subs %>% 
+  filter(QuestionName=="WR LL Eligibility-Medical" & Year==2018) %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(Medical=QuestionResults) %>% distinct(.keep_all=TRUE) %>% mutate(Medical="Yes")
+
+subsidies_O <- subs %>% 
+  filter(QuestionName=="WR LL Eligibility-Other" & Year==2018) %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(Other=QuestionResults) %>% distinct(.keep_all=TRUE) %>% mutate(Other="Yes")
+
+subsidies_HowFunded <- subs %>% 
+  filter(QuestionName=="WR LL Funded") %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(Disabled=QuestionResults) %>% distinct(.keep_all=TRUE)
+
+subsidies_BenefitForm <- subs %>% 
+  filter(QuestionName=="WR LL Benefit Form") %>% 
+  filter(QuestionResults!="-") %>%
+  select(PWSID,QuestionResults) %>%
+  rename(Disabled=QuestionResults) %>% distinct(.keep_all=TRUE)
+
+shutoff_pa<-left_join(b,shutoff_pa,by="PWSID")
+
+subs_b <- b %>% 
+  left_join(subsidies,by="PWSID") %>%
+  left_join(subsidies_LI,by="PWSID") %>%
+  left_join(subsidies_S,by="PWSID") %>%
+  left_join(subsidies_D,by="PWSID") %>%
+  left_join(subsidies_M,by="PWSID") %>%
+  left_join(subsidies_O,by="PWSID")
+
+subs_b$subsidy[which(subs_b$LowIncome=="Yes" | subs_b$Disabled=="Yes" | subs_b$Other=="Yes" | subs_b$Medical=="Yes" | subs_b$Seniors=="Yes")]<-"Yes"
+subs_b$LowIncome[which(subs_b$subsidy=="No")]<-"No"
+subs_b$Disabled[which(subs_b$subsidy=="No")]<-"No"
+subs_b$Medical[which(subs_b$subsidy=="No")]<-"No"
+subs_b$Seniors[which(subs_b$subsidy=="No")]<-"No"
+subs_b$Other[which(subs_b$subsidy=="No")]<-"No"
+
+subs_b$LowIncome[which(subs_b$subsidy=="Yes" & is.na(subs_b$LowIncome))]<-"No"
+subs_b$Disabled[which(subs_b$subsidy=="Yes" & is.na(subs_b$Disabled))]<-"No"
+subs_b$Medical[which(subs_b$subsidy=="Yes" & is.na(subs_b$Medical))]<-"No"
+subs_b$Seniors[which(subs_b$subsidy=="Yes" & is.na(subs_b$Seniors))]<-"No"
+subs_b$Other[which(subs_b$subsidy=="Yes" & is.na(subs_b$Other))]<-"No"
+
+```
 
@@
 
+sd<-lm(serious_violator_last_decade ~ Ownership + log(D_POPULATI),
+  data=filter(bv,D_POPULATI>3300)
+)
+summary(sd)
+
 ```

### 2020-09-20 paper.Rmd
update code

@@
 ## Profile of prevalence of private water utilities in CA (c. 2017))
+```{r ear}
+inv <- read_excel("Data/EAR/inventory.xlsx",
+                  sheet = "SDWISWaterSystemInventory071620") %>% filter(WaterSystemType=="C")
 
-There are ~ 2,800 Community Water Systems (defined as water systems serving at least 15 residential units on a year-round basis, generally applying to all municipal and institutional systems with residential components, but not commercial, school, and recreational facilities).
-Of these, 97.6% of the population is  served by the 653 systems that serve more than 3,300 people, while only 2.4% of the population is served by 2,173 small systems.
+inv$size_cat[which(inv$Pop<=500)] = "1. Very Small (<500)" 
+inv$size_cat[which(inv$Pop>500)] = "2. Small (501-3300)" 
+inv$size_cat[which(inv$Pop>3301)] = "3. Medium (3301-10,000)" 
+inv$size_cat[which(inv$Pop>10001)] = "4. Large (10,001- 100,000)" 
+inv$size_cat[which(inv$Pop>100000)] = "5. Very Large (>100,000)"
+
+size_summary <- inv %>% 
+  group_by(size_cat) %>%
+  add_tally() %>%
+  summarise(count=mean(n),pop=sum(Pop)/10^6) %>%
+  ungroup() %>%
+  mutate(perc_pop=100*pop/sum(pop))
+kable(size_summary, digits=1)
+
+# ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
+# ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
+# ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
+# ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
+# ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
+# ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
+# 
+# ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
+# rm(ear13,ear14,ear15,ear16,ear17,ear18)
+# save(ear,file="Data/EAR/ear.rds")
+load("Data/EAR/ear.rds")
+
+own <- ear %>% 
+    filter(QuestionName == "Population" | 
+                QuestionName == "Water System Classification" |
+                QuestionName == "Water System Ownership" |
+                QuestionName == "IsWholesaler") %>% 
+    mutate(QuestionID=paste0("q",QuestionID)) %>%
+    select(PWSID, Year, QuestionName,  QuestionResults) %>% 
+    distinct(PWSID, Year, QuestionName, .keep_all=TRUE) %>%
+    pivot_wider(names_from = QuestionName,
+                values_from = QuestionResults) %>%
+    group_by(PWSID) %>%
+    mutate(maxYear=max(Year)) %>%
+    ungroup() %>%
+    filter(Year == maxYear) %>%
+    filter(is.na(IsWholesaler)) %>%
+ #   mutate(Population = as.numeric(Population)) %>%
+    filter(`Water System Classification`=="Community Water System")
+
+own_inv <- left_join(inv,own,by="PWSID")
+
+```
+
+There are  2,917 Community Water Systems (defined as water systems serving at least 15 residential units on a year-round basis, generally applying to all municipal and institutional systems with residential components, but not commercial, school, and recreational facilities).
+Of these, 97.6% of the population is  served by the 658 systems that serve more than 3,300 people, while only 2.4% of the population is served by 2,259
+small systems.
 
@@
 
-A  caveat here is that this subset of utilties for the most part represents phenomenon in larger California utilities.
+A  caveat here is that this subset of utilities for the most part represents phenomenon in larger California utilities.
 Cost structures and dynamics affecting water pricing and privatization in rural areas may differ substantially from those in urban areas.

### 2020-09-20 paper.Rmd
update

@@
 
+questions <- c("Population",
+               "Water System Classification",
+               "Water System Ownership",
+               "IsWholesaler",
+               "WR Type",
+               "WR 6 HCF DW Charges",
+               "WR 12 HCF DW Charges",
+               "T Potable Total",
+               "WR SHUT_OFFS Once No Collect",
+               "WR SHUT_OFFS Once SF Total",
+               "WR SHUT_OFFS More Than Once MF Total",
+               "WR SHUT_OFFS More Than Once SF Total",
+               "WR SHUT_OFFS Once MF Total",
+               "WR LL Subsidies",
+               "WR LL Eligibility-Low Income",
+               "WR LL Eligibility-Disabled",
+               "WR LL Eligibility-Medical",
+               "WR LL Eligibility-Other",
+               "WR LL Eligibility-Senior",
+               "WR LL Benefit Form",
+               "WR LL Funded",
+               "DIST Probs SC BL Num",
+                "DIST Probs SC BL Investigated",
+                "DIST Probs SC BL Reported to District",
+                "DIST Probs SC BL Cause",
+                "DIST Probs Main BL Num",
+                "DIST Probs Main BL Investigated",
+                "DIST Probs Main BL Reported to District",
+                "DIST Probs Main BL Cause",
+                "DIST Probs WO Num",
+                "DIST Probs WO Investigated",
+                "DIST Probs WO Reported to District",
+                "DIST Probs WO Cause",
+                "DIST Probs BWO Num",
+                "DIST Probs BWO Investigated",
+                "DIST Probs BWO Reported to District",
+                "DIST Probs BWO Cause",
+                "DIST Probs Total Num",
+                "DIST Probs Total Investigated",
+                "DIST Probs Total Reported to District",
+                "DIST Probs Comments",
+                "TREATMENT GW Problems",
+                "TREATMENT SW Problems",
+                "SLI Lead Number",
+                "SLI Lead Total Length",
+                "SLI Fittings NOT lead pipe",
+                "SLI Fittings Lead pipe",
+                "SLI Total Lead Service Lines",
+                "SLI Cast Number",
+                "SLI Copper Number",
+                "SLI Ductile Number",
+                "SLI Fittings Unknown",
+                "SLI Galvanized Number",
+                "SLI HDPE Number",
+                "SLI Identify material 1",
+                "SLI Identify material 2",
+                "SLI Identify material 3",
+                "SLI Identify material 4",
+                "SLI Material 1 Number",
+                "SLI Material 2 Number",
+                "SLI Material 3 Number",
+                "SLI Material 4 Number",
+                "SLI Polybutylene Number",
+                "SLI Polyethylene Number",
+                "SLI Polyvinyl Number",
+                "SLI TA cement Number",
+                "SLI Total Lead Service Lines",
+                "SLI Total Service Connections",
+                "SLI Total Service Lines",
+                "SLI Unknown Number",
+                "SLI Unknown Total Length",
+                "DIST Infra Rehabilitation Comments",
+                "DIST Infra Rehabilitation Expenditure",
+                "DIST Infra Rehabilitation Not Tracked",
+                "DIST Infra Rehabilitation Num Leaks",
+                "DIST Infra Rehabilitation Portion",
+                "DIST Infra Rehabilitation Real Loss",
+                "DIST Infra renewal table unit",
+                "DIST Infra Repair Comments",
+                "DIST Infra Repair Expenditure",
+                "DIST Infra Repair Not Tracked",
+                "DIST Infra Repair Num Leaks",
+                "DIST Infra Repair Portion",
+                "DIST Infra Repair Real Loss",
+                "DIST Infra Replacement Comments",
+                "DIST Infra Replacement Expenditure",
+                "DIST Infra Replacement Not Tracked",
+                "DIST Infra Replacement Num Leaks",
+                "DIST Infra Replacement Portion",
+                "DIST Infra Replacement Real Loss",
+                "DIST Infra Total Comments",
+                "DIST Infra Total Expenditure",
+                "DIST Infra Total Not Tracked",
+                "DIST Infra Total Num Leaks",
+                "DIST Infra Total Portion",
+                "DIST Infra Total Real Loss",
+                "Cur Total Connections",
+                "Inactive Connections",
+                "Connections meter records kept",
+                "Connections SF with outdoor irrigation meters",
+                "Connections MF with outdoor irrigation meters",
+                "Connections customers with indoor submeters",
+                "Connections Comment",
+                "DIST Interv Booster Average",
+                "DIST Interv Booster Comments",
+                "DIST Interv Booster Expenditure",
+                "DIST Interv Booster Not Tracked",
+                "DIST Interv Booster Portion",
+                "DIST Interv Other Average",
+                "DIST Interv Other Comments",
+                "DIST Interv Other Expenditure",
+                "DIST Interv Other Not Tracked",
+                "DIST Interv Other Portion",
+                "DIST Interv Pressure R Average",
+                "DIST Interv Pressure R Comments",
+                "DIST Interv Pressure R Expenditure",
+                "DIST Interv Pressure R Not Tracked",
+                "DIST Interv Pressure R Portion",
+                "DIST Interv Pressure T Average",
+                "DIST Interv Pressure T Comments",
+                "DIST Interv Pressure T Expenditure",
+                "DIST Interv Pressure T Not Tracked",
+                "DIST Interv Pressure T Portion",
+                "DIST Interv Reduced Average",
+                "DIST Interv Reduced Comments",
+                "DIST Interv Reduced Expenditure",
+                "DIST Interv Reduced Not Tracked",
+                "DIST Interv Reduced Portion",
+                "DIST Interv Total Average",
+                "DIST Interv Total Comments",
+                "DIST Interv Total Expenditure",
+                "DIST Interv Total Not Tracked",
+                "DIST Interv Total Portion",
+                "DIST 5 Year Expenditure",
+                "DIST 5 Year Summary Comments",
+                "DIST 5 Year Total Leaks",
+                "DIST 5 Year Water Loss" 
+               )
+
 own <- ear %>% 
-    filter(QuestionName == "Population" | 
-                QuestionName == "Water System Classification" |
-                QuestionName == "Water System Ownership" |
-                QuestionName == "IsWholesaler") %>% 
-    mutate(QuestionID=paste0("q",QuestionID)) %>%
+    filter(QuestionName %in% questions) %>% 
+    group_by(PWSID,Year,QuestionName) %>%
+     mutate(maxYear=max(Year)) %>%
+    ungroup() %>%
+  #  filter(Year == maxYear) %>%
    select(PWSID, Year, QuestionName,  QuestionResults) %>% 
@@
                 values_from = QuestionResults) %>%
-    group_by(PWSID) %>%
-    mutate(maxYear=max(Year)) %>%
-    ungroup() %>%
-    filter(Year == maxYear) %>%
+
     filter(is.na(IsWholesaler)) %>%
@@
     filter(`Water System Classification`=="Community Water System")
-
-own_inv <- left_join(inv,own,by="PWSID")
 

### 2020-09-20 paper.Rmd
update

@@
     filter(`Water System Classification`=="Community Water System")
+
+save(own,"Data/own.rds")
+
+# census data
+census_api_key("b25f8b1b7bf10561c9cbc3a20a4d2572677f1f05")
+options(tigris_use_cache = TRUE)
+
+vars <- load_variables(2018, "acs5", cache = TRUE)
+
+census_vars <- c(median_income="B19013_001",
+                 pop_in_hh_with_income_assistance="B09010_002",
+                 pop_in_hh_with_no_income_assistance="B09010_008",
+                 hh_inc_5k="B19001_002",
+                 hh_inc_12.5k="B19001_003",
+                 hh_inc_17.5k="B19001_004",
+                 hh_inc_22.5k="B19001_005",
+                 hh_inc_27.5k="B19001_006",
+                 hh_inc_32.5k="B19001_007",
+                 hh_inc_37.5k="B19001_008",
+                 hh_inc_42.5k="B19001_009",
+                 hh_inc_47.5k="B19001_010",
+                 hh_inc_55k="B19001_011",
+                 hh_inc_67.5k="B19001_012",
+                 hh_inc_87.5k="B19001_013",
+                 hh_inc_112.5k="B19001_014",
+                 hh_inc_137.5k="B19001_015",
+                 hh_inc_175k="B19001_016",
+                 hh_inc_gr200k="B19001_017",
+                 hh_income_count = "B19001_001",
+                 hh_med_income_count = "B19019_001",
+                 hh_size_1_med_income = "B19019_002",
+                 hh_size_2_med_income = "B19019_003",
+                 hh_size_3_med_income = "B19019_004",
+                 hh_size_4_med_income = "B19019_005",
+                 hh_size_5_med_income = "B19019_006",
+                 hh_size_6_med_income = "B19019_007",
+                 hh_size_7plus_med_income = "B19019_008",
+                 avg_hh_size = "B25010_001",
+                 hh_inc_5k_rentcalc = 'B25122_002',
+                 hh_inc_15k_rentcalc = 'B25122_019',
+                 hh_inc_27.5k_rentcalc = 'B25122_036',
+                 hh_inc_42.5k_rentcalc = 'B25122_053',
+                 hh_inc_67.5k_rentcalc = 'B25122_070',
+                 hh_inc_87.5k_rentcalc = 'B25122_087',
+                 hh_inc_150k_rentcalc = 'B25122_104',
+                 hh_inc_5k_with_rent = 'B25122_004',
+                 hh_inc_5k_gross_rent_50 = 'B25122_004',
+                 hh_inc_5k_gross_rent_150 = 'B25122_005',
+                 hh_inc_5k_gross_rent_250 = 'B25122_006',
+                 hh_inc_5k_gross_rent_350 = 'B25122_007',
+                 hh_inc_5k_gross_rent_450 = 'B25122_008',
+                 hh_inc_5k_gross_rent_550 = 'B25122_009',
+                 hh_inc_5k_gross_rent_650 = 'B25122_010',
+                 hh_inc_5k_gross_rent_750 = 'B25122_011',
+                 hh_inc_5k_gross_rent_850 = 'B25122_012',
+                 hh_inc_5k_gross_rent_950 = 'B25122_013',
+                 hh_inc_5k_gross_rent_1100 = 'B25122_014',
+                 hh_inc_5k_gross_rent_1325 = 'B25122_015',
+                 hh_inc_5k_gross_rent_1750 = 'B25122_016',
+                 hh_inc_5k_gross_rent_2500 = 'B25122_017',
+                 hh_inc_15k_with_rent = 'B25122_020',
+                 hh_inc_15k_gross_rent_50 = 'B25122_021',
+                 hh_inc_15k_gross_rent_150 = 'B25122_022',
+                 hh_inc_15k_gross_rent_250 = 'B25122_023',
+                 hh_inc_15k_gross_rent_350 = 'B25122_024',
+                 hh_inc_15k_gross_rent_450 = 'B25122_025',
+                 hh_inc_15k_gross_rent_550 = 'B25122_026',
+                 hh_inc_15k_gross_rent_650 = 'B25122_027',
+                 hh_inc_15k_gross_rent_750 = 'B25122_028',
+                 hh_inc_15k_gross_rent_850 = 'B25122_029',
+                 hh_inc_15k_gross_rent_950 = 'B25122_030',
+                 hh_inc_15k_gross_rent_1100 = 'B25122_031',
+                 hh_inc_15k_gross_rent_1325 = 'B25122_032',
+                 hh_inc_15k_gross_rent_1750 = 'B25122_033',
+                 hh_inc_15k_gross_rent_2500 = 'B25122_034',         
+                 hh_inc_27.5k_with_rent = 'B25122_037',
+                 hh_inc_27.5k_gross_rent_50 = 'B25122_038',
+                 hh_inc_27.5k_gross_rent_150 = 'B25122_039',
+                 hh_inc_27.5k_gross_rent_250 = 'B25122_040',
+                 hh_inc_27.5k_gross_rent_350 = 'B25122_041',
+                 hh_inc_27.5k_gross_rent_450 = 'B25122_042',
+                 hh_inc_27.5k_gross_rent_550 = 'B25122_043',
+                 hh_inc_27.5k_gross_rent_650 = 'B25122_044',
+                 hh_inc_27.5k_gross_rent_750 = 'B25122_045',
+                 hh_inc_27.5k_gross_rent_850 = 'B25122_046',
+                 hh_inc_27.5k_gross_rent_950 = 'B25122_047',
+                 hh_inc_27.5k_gross_rent_1100 = 'B25122_048',
+                 hh_inc_27.5k_gross_rent_1325 = 'B25122_049',
+                 hh_inc_27.5k_gross_rent_1750 = 'B25122_050',
+                 hh_inc_27.5k_gross_rent_2500 = 'B25122_051',                  
+                 hh_inc_42.5k_with_rent = 'B25122_054',
+                 hh_inc_42.5k_gross_rent_50 = 'B25122_055',
+                 hh_inc_42.5k_gross_rent_150 = 'B25122_056',
+                 hh_inc_42.5k_gross_rent_250 = 'B25122_057',
+                 hh_inc_42.5k_gross_rent_350 = 'B25122_058',
+                 hh_inc_42.5k_gross_rent_450 = 'B25122_059',
+                 hh_inc_42.5k_gross_rent_550 = 'B25122_060',
+                 hh_inc_42.5k_gross_rent_650 = 'B25122_061',
+                 hh_inc_42.5k_gross_rent_750 = 'B25122_062',
+                 hh_inc_42.5k_gross_rent_850 = 'B25122_063',
+                 hh_inc_42.5k_gross_rent_950 = 'B25122_064',
+                 hh_inc_42.5k_gross_rent_1100 = 'B25122_065',
+                 hh_inc_42.5k_gross_rent_1325 = 'B25122_066',
+                 hh_inc_42.5k_gross_rent_1750 = 'B25122_067',
+                 hh_inc_42.5k_gross_rent_2500 = 'B25122_068',
+                 hh_inc_62.5k_with_rent = 'B25122_071',
+                 hh_inc_62.5k_gross_rent_50 = 'B25122_072',
+                 hh_inc_62.5k_gross_rent_150 = 'B25122_073',
+                 hh_inc_62.5k_gross_rent_250 = 'B25122_074',
+                 hh_inc_62.5k_gross_rent_350 = 'B25122_075',
+                 hh_inc_62.5k_gross_rent_450 = 'B25122_076',
+                 hh_inc_62.5k_gross_rent_550 = 'B25122_077',
+                 hh_inc_62.5k_gross_rent_650 = 'B25122_078',
+                 hh_inc_62.5k_gross_rent_750 = 'B25122_079',
+                 hh_inc_62.5k_gross_rent_850 = 'B25122_080',
+                 hh_inc_62.5k_gross_rent_950 = 'B25122_081',
+                 hh_inc_62.5k_gross_rent_1100 = 'B25122_082',
+                 hh_inc_62.5k_gross_rent_1325 = 'B25122_083',
+                 hh_inc_62.5k_gross_rent_1750 = 'B25122_084',
+                 hh_inc_62.5k_gross_rent_2500 = 'B25122_085', 
+                 hh_inc_82.5k_with_rent = 'B25122_088',
+                 hh_inc_82.5k_gross_rent_50 = 'B25122_089',
+                 hh_inc_82.5k_gross_rent_150 = 'B25122_090',
+                 hh_inc_82.5k_gross_rent_250 = 'B25122_091',
+                 hh_inc_82.5k_gross_rent_350 = 'B25122_092',
+                 hh_inc_82.5k_gross_rent_450 = 'B25122_093',
+                 hh_inc_82.5k_gross_rent_550 = 'B25122_094',
+                 hh_inc_82.5k_gross_rent_650 = 'B25122_095',
+                 hh_inc_82.5k_gross_rent_750 = 'B25122_096',
+                 hh_inc_82.5k_gross_rent_850 = 'B25122_097',
+                 hh_inc_82.5k_gross_rent_950 = 'B25122_098',
+                 hh_inc_82.5k_gross_rent_1100 = 'B25122_099',
+                 hh_inc_82.5k_gross_rent_1325 = 'B25122_100',
+                 hh_inc_82.5k_gross_rent_1750 = 'B25122_101',
+                 hh_inc_82.5k_gross_rent_2500 = 'B25122_102', 
+                 hh_inc_150k_with_rent = 'B25122_105',
+                 hh_inc_150k_gross_rent_50 = 'B25122_106',
+                 hh_inc_150k_gross_rent_150 = 'B25122_107',
+                 hh_inc_150k_gross_rent_250 = 'B25122_108',
+                 hh_inc_150k_gross_rent_350 = 'B25122_109',
+                 hh_inc_150k_gross_rent_450 = 'B25122_110',
+                 hh_inc_150k_gross_rent_550 = 'B25122_111',
+                 hh_inc_150k_gross_rent_650 = 'B25122_112',
+                 hh_inc_150k_gross_rent_750 = 'B25122_113',
+                 hh_inc_150k_gross_rent_850 = 'B25122_114',
+                 hh_inc_150k_gross_rent_950 = 'B25122_115',
+                 hh_inc_150k_gross_rent_1100 = 'B25122_116',
+                 hh_inc_150k_gross_rent_1325 = 'B25122_117',
+                 hh_inc_150k_gross_rent_1750 = 'B25122_118',
+                 hh_inc_150k_gross_rent_2500 = 'B25122_119' 
+                 )
+
+tracts <- get_acs(state = "CA", geography = "tract",
+                  variables = census_vars, geometry = TRUE)
+save(tracts,file="Data/census_tracts.rds")
+
+# census_vars2 <- c(median_income="B19013_001",
+#                  pop_in_hh_with_income_assistance="B09010_002",
+#                  pop_in_hh_with_no_income_assistance="B09010_008",
+#                  hh_inc_5k="B19001_002",
+#                  hh_inc_12.5k="B19001_003",
+#                  hh_inc_17.5k="B19001_004",
+#                  hh_inc_22.5k="B19001_005",
+#                  hh_inc_27.5k="B19001_006",
+#                  hh_inc_32.5k="B19001_007",
+#                  hh_inc_37.5k="B19001_008",
+#                  hh_inc_42.5k="B19001_009",
+#                  hh_inc_47.5k="B19001_010",
+#                  hh_inc_55k="B19001_011",
+#                  hh_inc_67.5k="B19001_012",
+#                  hh_inc_87.5k="B19001_013",
+#                  hh_inc_112.5k="B19001_014",
+#                  hh_inc_137.5k="B19001_015",
+#                  hh_inc_175k="B19001_016",
+#                  hh_inc_gr200k="B19001_017",
+#                  hh_income_count = "B19001_001")
+
+bg <- get_acs(state = "CA", geography = "block group",
+                  variables = census_vars, geometry = TRUE)
+
+save(bg,file="Data/census_blockgroups.rds")
+
+b.x <- st_buffer(st_transform(bounds,3310),dist=0)
+t.x <- tracts %>%
+  select(GEOID,geometry) %>%
+  st_transform(3310) %>%
+  st_buffer(dist=0) %>%
+  mutate(AreaTract = st_area(.))
+
+  st_buffer(st_transform(tracts,3310),dist=0)
+bg.x <- st_buffer(st_transform(bg,3310),dist=0)
+t.x$AreaTract_Total <- st_area(t.x)
+bg.x$AreaBlockGroup_Total <- st_area(bg.x)
+
+tracts.bounds <- st_intersection(b.x,t.x) 
+tracts.bounds$Area_TractPWS <- st_area(tracts.bounds)
+tracts.bounds$Percent_TractinPWS <- tracts.bounds$Area_TractPWS/tracts.bounds$AreaTract_Total
+
+bg.bounds <- st_intersection(bg.x,t.x) 
+bg.bounds$Area_BlockGroupPWS <- st_area(bg.bounds)
+bg.bounds$Percent_BlockGroupinPWS <- bg.bounds$Area_BlockGroupPWS/bg.bounds$AreaBlockGroup_Total
+
+### Find 20% and 50% Income quintile ranges by CWS (use block groups)
+
+### Find "disposable income" at 20% range
+
+##### 1. 
+##Take rent distribution within the range corresponding to the bracket that the 20% income quintile is.
+
+## Take median gross rent.
+## Calc "disoposable income" = 20% income/12 - median gross rent + water at 6hcf bill
+## Calc AR20_hh3 = 6hcf bill/disposable income
+## Cacl AR20_hh6 = 12hcf bill/ disposable incoem
+
+tracts.b<-tracts[bounds,]
+
+tracts.b %>%
+  ggplot(aes(fill = estimate)) +
+  geom_sf(color = NA) +
+  coord_sf(crs = 26911) +
+  scale_fill_viridis_c(option = "magma")
+
+## Assign tracts to pws
 

### 2020-09-21 paper.Rmd
update

@@
 
-# census_vars2 <- c(median_income="B19013_001",
-#                  pop_in_hh_with_income_assistance="B09010_002",
-#                  pop_in_hh_with_no_income_assistance="B09010_008",
-#                  hh_inc_5k="B19001_002",
-#                  hh_inc_12.5k="B19001_003",
-#                  hh_inc_17.5k="B19001_004",
-#                  hh_inc_22.5k="B19001_005",
-#                  hh_inc_27.5k="B19001_006",
-#                  hh_inc_32.5k="B19001_007",
-#                  hh_inc_37.5k="B19001_008",
-#                  hh_inc_42.5k="B19001_009",
-#                  hh_inc_47.5k="B19001_010",
-#                  hh_inc_55k="B19001_011",
-#                  hh_inc_67.5k="B19001_012",
-#                  hh_inc_87.5k="B19001_013",
-#                  hh_inc_112.5k="B19001_014",
-#                  hh_inc_137.5k="B19001_015",
-#                  hh_inc_175k="B19001_016",
-#                  hh_inc_gr200k="B19001_017",
-#                  hh_income_count = "B19001_001")
+census_vars2 <- c(median_income="B19013_001",
+                 pop_in_hh_with_income_assistance="B09010_002",
+                 pop_in_hh_with_no_income_assistance="B09010_008",
+                 hh_inc_5k="B19001_002",
+                 hh_inc_12.5k="B19001_003",
+                 hh_inc_17.5k="B19001_004",
+                 hh_inc_22.5k="B19001_005",
+                 hh_inc_27.5k="B19001_006",
+                 hh_inc_32.5k="B19001_007",
+                 hh_inc_37.5k="B19001_008",
+                 hh_inc_42.5k="B19001_009",
+                 hh_inc_47.5k="B19001_010",
+                 hh_inc_55k="B19001_011",
+                 hh_inc_67.5k="B19001_012",
+                 hh_inc_87.5k="B19001_013",
+                 hh_inc_112.5k="B19001_014",
+                 hh_inc_137.5k="B19001_015",
+                 hh_inc_175k="B19001_016",
+                 hh_inc_gr200k="B19001_017",
+                 hh_income_count = "B19001_001")
 
 bg <- get_acs(state = "CA", geography = "block group",
-                  variables = census_vars, geometry = TRUE)
+                  variables = census_vars2, geometry = TRUE)
 
@@
 b.x <- st_buffer(st_transform(bounds,3310),dist=0)
-t.x <- tracts %>%
-  select(GEOID,geometry) %>%
+t.x <- tracts %>% filter(variable=="median_income") %>%
+  select(GEOID,NAME,geometry) %>%
   st_transform(3310) %>%
@@
 
-  st_buffer(st_transform(tracts,3310),dist=0)
-bg.x <- st_buffer(st_transform(bg,3310),dist=0)
-t.x$AreaTract_Total <- st_area(t.x)
-bg.x$AreaBlockGroup_Total <- st_area(bg.x)
-
 tracts.bounds <- st_intersection(b.x,t.x) 
 tracts.bounds$Area_TractPWS <- st_area(tracts.bounds)
-tracts.bounds$Percent_TractinPWS <- tracts.bounds$Area_TractPWS/tracts.bounds$AreaTract_Total
+tracts.bounds$Percent_TractinPWS <- tracts.bounds$Area_TractPWS/tracts.bounds$AreaTract
 
@@
 ### Find 20% and 50% Income quintile ranges by CWS (use block groups)
+tract.income <- tracts %>% filter(variable %in% c('hh_inc_5k',
+                 'hh_inc_12.5k',
+                 'hh_inc_17.5k',
+                 'hh_inc_22.5k',
+                 'hh_inc_27.5k',
+                 'hh_inc_32.5k',
+                 'hh_inc_37.5k',
+                 'hh_inc_42.5k',
+                 'hh_inc_47.5k',
+                 'hh_inc_55k',
+                 'hh_inc_67.5k',
+                 'hh_inc_87.5k',
+                 'hh_inc_112.5k',
+                 'hh_inc_137.5k',
+                 'hh_inc_175k',
+                 'hh_inc_gr200k')) %>%
+  st_drop_geometry() %>% group_by(GEOID) %>%
+  mutate(cum_inc_sum = cumsum(estimate),
+         percentile_inc = 100*cum_inc_sum/sum(estimate))
 

### 2020-09-21 paper.Rmd
update

@@
 library(stargazer)
+library(readxl)
 ```
@@
 
+bounds <- st_read("Data/boundaries.geojson") %>% 
+  filter(D_PWS_FED_=="C") %>% 
+  select(PWSID,D_POPULATI,SVC_CONNEC)
+
 b.x <- st_buffer(st_transform(bounds,3310),dist=0)
@@
 
+bg.x <- bg %>% filter(variable=="median_income") %>%
+  select(GEOID,NAME,geometry) %>%
+  st_transform(3310) %>%
+  st_buffer(dist=0) %>%
+  mutate(AreaBlockGroup = st_area(.))
+
 tracts.bounds <- st_intersection(b.x,t.x) 
@@
 tracts.bounds$Percent_TractinPWS <- tracts.bounds$Area_TractPWS/tracts.bounds$AreaTract
+save(tracts.bounds,file="Data/tracts_bounds.rds")
 
-bg.bounds <- st_intersection(bg.x,t.x) 
+bg.bounds <- st_intersection(b.x,bg.x) 
 bg.bounds$Area_BlockGroupPWS <- st_area(bg.bounds)
-bg.bounds$Percent_BlockGroupinPWS <- bg.bounds$Area_BlockGroupPWS/bg.bounds$AreaBlockGroup_Total
+bg.bounds$Percent_BlockGroupinPWS <- bg.bounds$Area_BlockGroupPWS/bg.bounds$AreaBlockGroup
+save(bg.bounds,file="Data/block_groups_bounds.rds")
 
 ### Find 20% and 50% Income quintile ranges by CWS (use block groups)
-tract.income <- tracts %>% filter(variable %in% c('hh_inc_5k',
+pws.income <- bg.bounds %>% st_drop_geometry() %>%
+  left_join(bg,by="GEOID") %>%
+  filter(variable %in% c('hh_inc_5k',
                  'hh_inc_12.5k',
@@
                  'hh_inc_gr200k')) %>%
-  st_drop_geometry() %>% group_by(GEOID) %>%
+  group_by(PWSID,variable) %>%
+  summarise(estimate=sum(estimate*Percent_BlockGroupinPWS)) %>% 
+  ungroup() %>%
+  mutate(var2=case_when(variable=="hh_inc_5k" ~ 1,
+                 variable=="hh_inc_12.5k" ~ 2,
+                 variable=="hh_inc_17.5k" ~ 3,
+                 variable=="hh_inc_22.5k" ~ 4,
+                 variable=="hh_inc_27.5k" ~ 5,
+                 variable=="hh_inc_32.5k" ~ 6,
+                 variable=="hh_inc_37.5k" ~ 7,
+                 variable=="hh_inc_42.5k" ~ 8,
+                 variable=="hh_inc_47.5k" ~ 9,
+                 variable=="hh_inc_55k" ~ 10,
+                 variable=="hh_inc_67.5k" ~ 11,
+                 variable=="hh_inc_87.5k" ~ 12,
+                 variable=="hh_inc_112.5k" ~ 13,
+                 variable=="hh_inc_137.5k" ~ 14,
+                 variable=="hh_inc_175k" ~ 15,
+                 variable=="hh_inc_gr200k" ~ 16,)) %>% 
+  arrange(PWSID,var2) %>% 
+  group_by(PWSID) %>%
   mutate(cum_inc_sum = cumsum(estimate),
-         percentile_inc = 100*cum_inc_sum/sum(estimate))
+         percentile_inc = 100*cum_inc_sum/sum(estimate)) %>%
+  ungroup() %>%
+  mutate(perc_dist20=abs(as.numeric(percentile_inc)-20),
+         perc_dist50=abs(as.numeric(percentile_inc)-50)) %>%
+  group_by(PWSID) %>%
+  filter(perc_dist20==min(perc_dist20) | 
+         perc_dist50==min(perc_dist50)) %>%
+  mutate(percentile=case_when(var2==min(var2)~"20th Percentile",
+                              var2==max(var2)~"50th Percentile"),
+         income_perc=case_when(var2==1 ~ 5000,
+                               var2==2 ~ 12500,
+                               var2==3 ~ 17500,
+                               var2==4 ~ 22500,
+                               var2==5 ~ 27500,
+                               var2==6 ~ 32500,
+                               var2==7 ~ 37500,
+                               var2==8 ~ 42500,
+                               var2==9 ~ 47500,
+                               var2==10 ~ 55000,
+                               var2==11 ~ 67500,
+                               var2==12 ~ 87500,
+                               var2==13 ~ 112500,
+                               var2==14 ~ 137500,
+                               var2==15 ~ 175000,
+                               var2==16 ~ 225000)
+         ) %>%
+  pivot_wider(id_cols=PWSID,names_from=percentile,values_from=income_perc)
+
+pws.income2 <-unnest(pws.income, cols = c(`20th Percentile`, `50th Percentile`, `NA`)) %>%
+  select(PWSID,`20th Percentile`,`50th Percentile`) %>% distinct(.keep_all=TRUE)
+
+save(pws.income2,file="Data/pws_income2.rds")
+
+load("Data/own.rds")
 
 ### Find "disposable income" at 20% range
+
+pws.rent <-  bg.bounds %>% 
+  st_drop_geometry() %>%
+  left_join(tracts,by="GEOID") %>%
+  filter(variable=)
 
@@
 94% of the population served by private utilities are served by utlities with more than 3,300 people.
-```{r boundaries, echo=FALSE,results='asis',message=FALSE, warning=FALSE}
-b <- st_read("Data/boundaries.geojson") %>% filter(D_PWS_FED_=="C") 
-b$size_cat= NA
-
-b$size_cat[which(b$D_POPULATI<=500)] = "1. Very Small (<500)" 
-b$size_cat[which(b$D_POPULATI>500)] = "2. Small (501-3300)" 
-b$size_cat[which(b$D_POPULATI>3301)] = "3. Medium (3301-10,000)" 
-b$size_cat[which(b$D_POPULATI>10001)] = "4. Large (10,001- 100,000)" 
-b$size_cat[which(b$D_POPULATI>100000)] = "5. Very Large (>100,000)"
-
-b$Ownership <-"Public"
-b$Ownership[which(b$OWNER_TYPE=="P")]<-"Private"
-
-size_summary <- b %>% group_by(size_cat)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry()
-
-size_summary$perc_pop <- 100*size_summary$pop/sum(size_summary$pop)
-size_summary <- arrange(size_summary,size_cat)
-kable(size_summary, caption= "Size and population distribution of CA Water Utilities",digits=1)
-
-private <- b%>%group_by(Ownership)%>%add_tally()%>%summarize(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() %>% ungroup()%>%mutate(perc_pop=100*pop/sum(pop))
-kable(private, caption= "Population served by public and private utilities in CA",digits=1)
-
-size_summary.p <- b %>% group_by(size_cat,Ownership)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry()
-
-size_summary.p$perc_pop <- 100*size_summary.p$pop/sum(size_summary.p$pop)
-size_summary.p <- arrange(size_summary.p,size_cat,Ownership)
-kable(size_summary.p, caption= "Size and population distribution and ownership of CA Water Utilities",digits=1)
-
-```
+<!-- # ```{r boundaries, echo=FALSE,results='asis',message=FALSE, warning=FALSE} -->
+<!-- # b <- st_read("Data/boundaries.geojson") %>% filter(D_PWS_FED_=="C")  -->
+<!-- # b$size_cat= NA -->
+<!-- #  -->
+<!-- # b$size_cat[which(b$D_POPULATI<=500)] = "1. Very Small (<500)"  -->
+<!-- # b$size_cat[which(b$D_POPULATI>500)] = "2. Small (501-3300)"  -->
+<!-- # b$size_cat[which(b$D_POPULATI>3301)] = "3. Medium (3301-10,000)"  -->
+<!-- # b$size_cat[which(b$D_POPULATI>10001)] = "4. Large (10,001- 100,000)"  -->
+<!-- # b$size_cat[which(b$D_POPULATI>100000)] = "5. Very Large (>100,000)"  -->
+<!-- #  -->
+<!-- # b$Ownership <-"Public" -->
+<!-- # b$Ownership[which(b$OWNER_TYPE=="P")]<-"Private" -->
+<!-- #  -->
+<!-- # size_summary <- b %>% group_by(size_cat)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() -->
+<!-- #  -->
+<!-- # size_summary$perc_pop <- 100*size_summary$pop/sum(size_summary$pop) -->
+<!-- # size_summary <- arrange(size_summary,size_cat) -->
+<!-- # kable(size_summary, caption= "Size and population distribution of CA Water Utilities",digits=1) -->
+<!-- #  -->
+<!-- # private <- b%>%group_by(Ownership)%>%add_tally()%>%summarize(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() %>% ungroup()%>%mutate(perc_pop=100*pop/sum(pop)) -->
+<!-- # kable(private, caption= "Population served by public and private utilities in CA",digits=1) -->
+<!-- #  -->
+<!-- # size_summary.p <- b %>% group_by(size_cat,Ownership)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() -->
+<!-- #  -->
+<!-- # size_summary.p$perc_pop <- 100*size_summary.p$pop/sum(size_summary.p$pop) -->
+<!-- # size_summary.p <- arrange(size_summary.p,size_cat,Ownership) -->
+<!-- # kable(size_summary.p, caption= "Size and population distribution and ownership of CA Water Utilities",digits=1) -->
+<!-- #  -->
+<!-- #  -->
+<!-- # ``` -->
 
@@
 rates <- read_csv("Data/affordability.csv")
+rates <- left_join(rates,pws.income2,by=c("pwsid"="PWSID"))
+rates$AR20 <- rates$total_bill_by_hhsize/rates$`20th Percentile`
+rates$MHI <- rates$total_bill_by_hhsize/rates$`50th Percentile`
+
 priv <- select(b,PWSID,Ownership,D_POPULATI)
@@
 rates$Private[which(rates$Ownership=="Private")]<-1
+ggplot(rates, aes(x=as.factor(hh_sizes), y=total_bill_by_hhsize, colour=Ownership)) +
+  geom_boxplot() 
 ggplot(rates, aes(x=as.factor(hh_sizes), y=total_bill_by_hhsize, colour=Ownership)) +
@@
 - Are richer places having higher prices?
+
+```{r richer higher}
+prices <- select(own,
+                 PWSID,
+                 Year,
+                 `Water System Classification`,
+                 `Water System Ownership`,
+                 `WR Type`,
+                 `WR 6 HCF DW Charges`,
+                 `WR 12 HCF DW Charges`) %>% filter(Year==2017 & `WR Type`!="-" & !is.na(`WR Type`))
+prices.income <- inv %>% left_join(prices,by="PWSID") %>% left_join(pws.income2,by="PWSID")
+```
+
 - Are private utiltiies tending to be in richer or poorer places than public utilities?
+
+```{r private richer}
+
+```
 

### 2020-09-21 Code/affordability_metrics.R

@@ -24,11 +24,11 @@ hh_sizes<-c(1:7) #household sizes in census go from 1 to 7
 gpcd<-50 # set gpcd for affordability threshold
 
 d<-crossing(d,hh_sizes)
-d%<>%mutate(reasonable_consumption_kgal=(gpcd*hh_sizes*365/12)/1000)
+d<- d%>%mutate(reasonable_consumption_kgal=(gpcd*hh_sizes*365/12)/1000)
 d$tier_0<-as.numeric(d$tier_0)
 
 #clean up water rates a bit, more, need to remove NAs from tier widths and tiers
-d%<>%mutate(tier_0 = case_when(is.na(tier_0) ~ 0,
+d<- d%>%mutate(tier_0 = case_when(is.na(tier_0) ~ 0,
                                TRUE ~ tier_0),
             tier_1 = case_when(is.na(tier_1) ~ 10^5,
                                TRUE ~ tier_1),
@@ -71,7 +71,7 @@ d%<>%mutate(tier_0 = case_when(is.na(tier_0) ~ 0,
 
 #function takes in summary data frame, and a vector of volumes, returns data frame with revenue for blocks and total water bill
 volRevCalc<-function(data,vol){
-  data%<>%mutate(
+  data<-data%>%mutate(
  # rev1=tier_1_price*pmax(pmin(vol,tier_0),0),
   rev1=tier_1_price*pmax((pmin(vol-tier_0,tier_1-tier_0)),0),
   rev2=tier_2_price*pmax((pmin(vol-tier_1,tier_2-tier_1)),0),
@@ -90,7 +90,7 @@ return(total_bill)
   
 }
 
-d%<>%mutate(total_bill_by_hhsize=volRevCalc(d,d$reasonable_consumption_kgal),
+d<-S%>%mutate(total_bill_by_hhsize=volRevCalc(d,d$reasonable_consumption_kgal),
                                             mean_bill=volRevCalc(d,d$usage_ccf*0.748))
 
 d$approximate_median_income[which(d$median_income_category=="$150,000 to $199,999")]<-175000
@@ -98,14 +98,14 @@ d$approximate_median_income[which(d$median_income_category=="$150,000 to $199,99
 ##### AFFORDABILITY METRIC 1: % Median HH Income
 # The most commonly used affordability metric is
 # the average water bill as a % of median area HH income (let's call perc_MHI)
[…]
